---
title: "Flocking and Foraging"
author: "Cody M Kent"
date: "February 7, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Flocking and Foraging Lab Data Analyis
##Introduction

1) This lab document goes with the provided R script that you will be working with.
2) You should have a second csv file titled FlockingAndForaging_Data.csv saved in the data folder.
3) This data file should already be cleaned and ready to go.
4) You should have a file titled FlockingAndForagingScript_Student.R that contains most of the script for today's analysis.

A final note before we begin. You will notice that the provided script is a bit sparse today. You should be able to use what you have learned to figure it out. Feel free to poach from your own code should you get stuck. That is what we do all of the time in real life

##Loading and examining data
First lets load in our that we collected in lab data file and check it for anything that doesn't make sense. Use the same commands we used before, including `head`, `tail`,`str`, and `summary`.

```{r load data 1, eval=FALSE}
Data<-  #load in the data
```
```{r load data 2, echo=FALSE}
Data<-read.csv('Data/FlockingAndForaging_Data.csv') #use the read.csv() file to load in the data
```

What structure is the data in (i.e. numeric, factor, logical)? Why is it like this, and how might it cause us some problems in the future?

If everything looks good then lets go ahead and organize the data. For our analyisis today we will want to have all of the data seperated by trial number. To do this use the `subset.data.frame()` function. Don't worry about trial 5. We won't be analyzing it today. The R I have placed infront of each object is just to denote that this is the raw data, as opposed to the summary data we wil calculate in the next step 

```{r devide by trial 1, eval=FALSE}
RTrial1<-
RTrial2<-
RTrial3<-
RTrial4<-
```

```{r devide by trial 2, echo=FALSE}
RTrial1<-subset.data.frame(Data,Data$Trial==1)
RTrial2<-subset.data.frame(Data,Data$Trial==2)
RTrial3<-subset.data.frame(Data,Data$Trial==3)
RTrial4<-subset.data.frame(Data,Data$Trial==4)
```

We also don't need the first two columns of data (Lab and Trail) any more, so lets get rid of them

```{r remove extra columns}
RTrial1<-RTrial1[,-c(1:2)] #remove first two columns
RTrial2<-RTrial2[,-c(1:2)]
RTrial3<-RTrial3[,-c(1:2)]
RTrial4<-RTrial4[,-c(1:2)]
```

Now remember that these are currently factors. We can't do things like calculate means with factors, for that we need numeric data. To do that, we will need to coerce this data into numeric. However, we also need to be able to tell who was killed by a hawk, so we will also keep these first data frames for that analysis. The code here is a bit ugly, so take some time now to break it down until you understand what is going on

```{r change to numeric}
RTrial1N<-as.data.frame(matrix(nrow=nrow(RTrial1),ncol=ncol(RTrial1),data=as.numeric(as.matrix(RTrial1))));colnames(RTrial1N)<-names(RTrial1)
RTrial2N<-as.data.frame(matrix(nrow=nrow(RTrial2),ncol=ncol(RTrial2),data=as.numeric(as.matrix(RTrial2))));colnames(RTrial2N)<-names(RTrial2)
RTrial3N<-as.data.frame(matrix(nrow=nrow(RTrial3),ncol=ncol(RTrial3),data=as.numeric(as.matrix(RTrial3))));colnames(RTrial3N)<-names(RTrial3)
RTrial4N<-as.data.frame(matrix(nrow=nrow(RTrial4),ncol=ncol(RTrial4),data=as.numeric(as.matrix(RTrial4))));colnames(RTrial4N)<-names(RTrial4)
```
##Calculating summary statistics

Next we will calculate a series of summary statistics for each day of each trial. The statsitcs you need to calculate are as follows: Mean time, standard deviation, total number of individuals, number starved, number depredated, number survived, percent starved, and percent killed. The code to do this for trial 1 is outlined for you. You will need to repeat it for the other three trials before moving on. 

First, we will calculate several of these values using our numeric data frames
```{r summary stats 1, eval=FALSE}
Trial1<-data.frame(MeanTime=apply(RTrial1N,2,mean,na.rm=T)) #This line will calculate the mean foraging time accross each column
#use the same format as above to calculate the standard deviation.
Trial1$SD<-sqrt(apply())   # remember standard deviation in the sqrt() of the var()
Trial1$N<-nrow() #finish the line to calculate the number of individuals
```

The next couple summary statistics we will get from our factor data frames
```{r summary stats 2,eval=F}
Trial1$NStarved<-colSums(RTrial1==60) #calculate the number in each column that equaled 60
Trial1$NDepredated<-   #Use the same format to calcualte number depredated
```

Finally, the last few summary stats we will calculate from some of the columns we have already created in `Trial1`
```{r summary stats 3,eval=F}
Trial1$NSurvived<- #use the colmuns of Trial 1 we already calculated 
Trial1$PStarved<- #caluclate percent starved
Trial1$PDepredated<- #caluclate percent starved
Trial1$PSurvived<-  #caclulate percent survived
```

```{r summary stats final, echo=FALSE}
Trial1<-data.frame(MeanTime=apply(RTrial1N,2,mean,na.rm=T)) #This line will calculate the mean foraging time accross each column
#use the same format as above to calculate the standard deviation
Trial1$SD<-sqrt(apply(RTrial1N,2,var,na.rm=T))   # remember standard deviation in the sqrt() of the var()
Trial1$N<-nrow(RTrial1N) #finish the line to calculate the number of individuals
#for the other calculations we need to use the first data set that had factors
Trial1$NStarved<-colSums(RTrial1=='60') #calculate the number in each column that equaled 60
Trial1$NDepredated<-colSums(RTrial1=='h')   #Use the same format to calcualte number depredated
#use the colmuns of Trial 1 we already calculated to get the next few
Trial1$NSurvived<-Trial1$N-Trial1$NStarved-Trial1$NDepredated #use the colmuns of Trial 1 we already calculated 
Trial1$PStarved<-Trial1$NStarved/Trial1$N #caluclate percent starved
Trial1$PDepredated<-Trial1$NDepredated/Trial1$N #caluclate percent starved
Trial1$PSurvived<-Trial1$NSurvived/Trial1$N #caclulate percent survived
```

Now make sure to take a look at your data frame you have now made using `summary()` and `str()` to make sure it looks correct. Assuming it does, repeat the calculaions for the other three trials. 
```{r summary stats other trials, echo=FALSE}
Trial2<-data.frame(MeanTime=apply(RTrial2N,2,mean,na.rm=T)) #This line will calculate the mean foraging time accross each column
#use the same format as above to calculate the standard deviation
Trial2$SD<-sqrt(apply(RTrial2N,2,var,na.rm=T))   # remember standard deviation in the sqrt() of the var()
Trial2$N<-nrow(RTrial2N) #finish the line to calculate the number of individuals
#for the other calculations we need to use the first data set that had factors
Trial2$NStarved<-colSums(RTrial2=='60') #calculate the number in each column that equaled 60
Trial2$NDepredated<-colSums(RTrial2=='h')   #Use the same format to calcualte number depredated
#use the colmuns of Trial 2 we already calculated to get the next few
Trial2$NSurvived<-Trial2$N-Trial2$NStarved-Trial2$NDepredated #use the colmuns of Trial 2 we already calculated 
Trial2$PStarved<-Trial2$NStarved/Trial2$N #caluclate percent starved
Trial2$PDepredated<-Trial2$NDepredated/Trial2$N #caluclate percent starved
Trial2$PSurvived<-Trial2$NSurvived/Trial2$N #caclulate percent survived

Trial3<-data.frame(MeanTime=apply(RTrial3N,2,mean,na.rm=T)) #This line will calculate the mean foraging time accross each column
#use the same format as above to calculate the standard deviation
Trial3$SD<-sqrt(apply(RTrial3N,2,var,na.rm=T))   # remember standard deviation in the sqrt() of the var()
Trial3$N<-nrow(RTrial3N) #finish the line to calculate the number of individuals
#for the other calculations we need to use the first data set that had factors
Trial3$NStarved<-colSums(RTrial3=='60') #calculate the number in each column that equaled 60
Trial3$NDepredated<-colSums(RTrial3=='h')   #Use the same format to calcualte number depredated
#use the colmuns of Trial 3 we already calculated to get the next few
Trial3$NSurvived<-Trial3$N-Trial3$NStarved-Trial3$NDepredated #use the colmuns of Trial 3 we already calculated 
Trial3$PStarved<-Trial3$NStarved/Trial3$N #caluclate percent starved
Trial3$PDepredated<-Trial3$NDepredated/Trial3$N #caluclate percent starved
Trial3$PSurvived<-Trial3$NSurvived/Trial3$N #caclulate percent survived

Trial4<-data.frame(MeanTime=apply(RTrial4N,2,mean,na.rm=T)) #This line will calculate the mean foraging time accross each column
#use the same format as above to calculate the standard deviation
Trial4$SD<-sqrt(apply(RTrial4N,2,var,na.rm=T))   # remember standard deviation in the sqrt() of the var()
Trial4$N<-nrow(RTrial4N) #finish the line to calculate the number of individuals
#for the other calculations we need to use the first data set that had factors
Trial4$NStarved<-colSums(RTrial4=='60') #calculate the number in each column that equaled 60
Trial4$NDepredated<-colSums(RTrial4=='h')   #Use the same format to calcualte number depredated
#use the colmuns of Trial 4 we already calculated to get the next few
Trial4$NSurvived<-Trial4$N-Trial4$NStarved-Trial4$NDepredated #use the colmuns of Trial 4 we already calculated 
Trial4$PStarved<-Trial4$NStarved/Trial4$N #caluclate percent starved
Trial4$PDepredated<-Trial4$NDepredated/Trial4$N #caluclate percent starved
Trial4$PSurvived<-Trial4$NSurvived/Trial4$N #caclulate percent survived
```

##Bin data into periods
As we have discussed before, it is often hard to seperate understand statics like chi-square or t-test results if you have too many categories. To help with this, we will only be using three days: one early (Day 3), one in the middle (Day 10), and one at the end (Day 17)


Now use this function to change each trial into a properly binned version.
```{r bin trials 1,eval=F}

#Grab only the days you need
Trial1<-Trial1[]
Trial2<-
Trial3<-
Trial4<-
  
#Also, lets add a 
Trial1$Day=c(3,10,17)
Trial2$Day=c(3,10,17)
Trial3$Day=c(3,10,17)
Trial4$Day=c(3,10,17)
```

```{r bin trials 2,echo=F}

#Now use the function to do the calculations for each trial.
Trial1<-Trial1[c(3,10,17),]
Trial2<-Trial2[c(3,10,17),]
Trial3<-Trial3[c(3,10,17),]
Trial4<-Trial4[c(3,10,17),]
#Also, lets add a 
Trial1$Day=c(3,10,17)
Trial2$Day=c(3,10,17)
Trial3$Day=c(3,10,17)
Trial4$Day=c(3,10,17)
```

Make sure to look at each trial and be sure it still look good and make sense. If they do, we can move on.

##Asignment
For the rest of the lab, you will be placed into groups. Each group will be assigned a question. It is up to your group to figure out how to analyze the data. Be sure to check in with me after you have some idea of how to analyze your question to make sure you are doing it right. At the end we will all come together and share our results. Also, I will have you send me your code for your question and compile them to be destributed to the class so everyone can see what you have done.

For your question, you will first need to figure out:
1) What trials will you use?
2) What variables will you test?
3) What statistical test will you use to compare them?
4) What figure can I make to show the results?
5) Make a results statement for your findings. 

Question 1 has been done for you. 
###Question 1
What was the effect of depleting resources on solitary foraging time?

1) For this question we are only interested in solitary foraging time without predation (Trial 1).
2) We will look at average foraging time
3) Because we are comparing two means, we will use a t-test
4) We will make a figure showing foraging time changing across time periods.
```{r Quesiton 1 t-test}
TTest<-function(m1,m2,SD1,SD2,N1,N2){
  t<-abs((m1-m2)/sqrt(SD1^2/N2+SD2^2/N1))
  df<-N1+N2-2
  p<-2*pt(t,df,lower.tail=F)
  Res<-data.frame(t,df,p)
  return(Res)
}  #Make a funciton to run the t-test

#grab the mean, SD, and N from trial 1. Use periods 1 and 2.
P1P2<-TTest(Trial1$MeanTime[1],Trial1$MeanTime[2],Trial1$SD[1],Trial1$SD[2],Trial1$N[1],Trial1$N[2]) 
P1P3<-TTest(Trial1$MeanTime[1],Trial1$MeanTime[3],Trial1$SD[1],Trial1$SD[3],Trial1$N[1],Trial1$N[3])
P2P3<-TTest(Trial1$MeanTime[2],Trial1$MeanTime[3],Trial1$SD[2],Trial1$SD[3],Trial1$N[2],Trial1$N[3])
P1P2;P1P3;P2P3 #show the results
```

We can see that they are all significantly different. Lets make a figure to help show how

```{r Question 1 figure}
plot(MeanTime~Day,data=Trial1,type='o',
     xlab='Day',ylab='Mean Foraging Time')
```

Then make a results statement.

###Question 2
What is the effect of competition in foraging time

```{r Quesiton 2 worked, echo=F,eval=F}
##Here we will compare trial 1 to trial 2 using t-tests on foraging time for each of the three time periods

P1<-TTest(Trial1$MeanTime[1],Trial2$MeanTime[1],Trial1$SD[1],Trial2$SD[1],Trial1$N[1],Trial2$N[1])
P2<-TTest(Trial1$MeanTime[2],Trial2$MeanTime[2],Trial1$SD[2],Trial2$SD[2],Trial1$N[2],Trial2$N[2])
P3<-TTest(Trial1$MeanTime[3],Trial2$MeanTime[3],Trial1$SD[3],Trial2$SD[3],Trial1$N[3],Trial2$N[3])


##And make a plot o the two
plot(MeanTime~Day,data=Trial1,type='o',
     xlab='Day',ylab='Mean Foraging Time',ylim=c(0,60))
points(MeanTime~Day,data=Trial2,type='o',col='red')
legend('topleft',legend=c('Trial1','Trial2'),lty=1,pch=1,col=c('black','red'))
```

###Question 3
What was the effect of predation on forager survivorship?

```{r Quesiton 3 worked, echo=F,eval=F}
##Here we will compare trial 1 and 3 with a chi-squared test
Obs<-data.frame(Trial1$NSurvived,Trial3$NSurvived)
Chi<-chisq.test(Obs)

##And make a plot
plot(PSurvived~Day,data=Trial1,type='o',
     xlab='Day',ylab='Mean Foraging Time',ylim=c(0,1))
points(PSurvived~Day,data=Trial3,type='o',col='red')
legend('bottomleft',legend=c('Trial1','Trial3'),lty=1,pch=1,col=c('black','red'))
```

###Question 4
Is flocking beneficial in terms of predation?

```{r Quesiton 4 worked, echo=F,eval=F}
##Here we will compare trial 3 and 4 with a chi-squared test
Obs<-data.frame(Trial3$NDepredated,Trial4$NDepredated)
Chi<-chisq.test(Obs)

##And make a plot
plot(PDepredated~Day,data=Trial3,type='o',
     xlab='Day',ylab='Mean Foraging Time',ylim=c(0,1))
points(PDepredated~Day,data=Trial4,type='o',col='red')
legend('topleft',legend=c('Trial3','Trial4'),lty=1,pch=1,col=c('black','red'))
```

