---
title: "Flocking and Foraging"
author: "Cody M Kent"
date: "February 7, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Flocking and Foraging Lab Data Analyis
##Introduction

1) This lab document goes with the provided R script that you will be working with.
2) You should have a second csv file titled FlockingAndForaging_Data.csv saved in the data folder.
3) This data file should already be cleaned and ready to go.
4) You should have a file titled FlockingAndForagingScript_Student.R that contains most of the script for today's analysis.

A final note before we begin. You will notice that the provided script is a bit sparse today. You should be able to use what you have learned to figure it out. Feel free to poach from your own code should you get stuck. That is what we do all of the time in real life

##Loading and examining data
First lets load in our that we collected in lab data file and check it for anything that doesn't make sense. Use the same commands we used before, including `head`, `tail`,`str`, and `summary`.

```{r load data 1, eval=FALSE}
Data<-  #load in the data
```
```{r load data 2, echo=FALSE}
Data<-read.csv('Data/FlockingAndForaging_Data.csv') #use the read.csv() file to load in the data
```

What structure is the data in (i.e. numeric, factor, logical)?


##Calculating summary statistics

Next we will calculate a series of summary statistics for each day of each trial. First we will find the average foraging time and its standard deviation, the total number of individuals participating, and the number starved or killed. To calculate the first round of these we will use something called piping. Here we will use to packages `dplyr` and `tidyr`. The special operation `%>%` tells R that we are going to pipe. Take some time to understand how this code works.


```{r summary stats 1, eval=FALSE}
library(tidyr);library(dplyr) #load in the packages we need to pipe
SummaryData<-Data %>% #make an object called SummaryData by using the object Data
  group_by(Trial,Day) %>% #group data by day, within trial
  summarise(MeanForagingTime=mean(ForagingTime,na.rm=T), #cacluate the average foraging time
            StandardDeviation=sqrt(var(ForagingTime,na.rm=T)), #standard deviation of foraging time
            N=length(unique(Group)), #number of individuals
            NStarved=sum(CauseOfDeath=='s'), #number starved
            NDepredated=sum(CauseOfDeath=='h' | CauseOfDeath=='g')) #number depredated
SummaryData<-as.data.frame(SummaryData) #Turn back into a data frame
```

Now you also need to calculate the number survived, percent starved, percent depredated, and percent survived. 

```{r summary stats 2, eval=F}
SummaryData$NSurvived<-
SummaryData$PStarved<-
SummaryData$PDepredated<-
SummaryData$PSurvived<-
```

```{r summary stats for real, echo=FALSE}
library(tidyr);library(dplyr)
SummaryData<-Data %>%
  group_by(Trial,Day) %>%
  summarise(MeanForagingTime=mean(ForagingTime,na.rm=T),
            StandardDeviation=sqrt(var(ForagingTime,na.rm=T)),
            N=length(unique(Group)),
            NStarved=sum(CauseOfDeath=='s'),
            NDepredated=sum(CauseOfDeath=='h' | CauseOfDeath=='g'))
SummaryData$NSurvived<-SummaryData$N-SummaryData$NStarved-SummaryData$NDepredated
SummaryData$PStarved<-SummaryData$NStarved/SummaryData$N
SummaryData$PDepredated<-SummaryData$NDepredated/SummaryData$N
SummaryData$PSurvived<-SummaryData$NSurvived/SummaryData$N
SummaryData<-as.data.frame(SummaryData)
```

Now make sure to take a look at your data frame you have now made using `summary()` and `str()` to make sure it looks correct.


##Organizing the data
###Limmiting the days of the data
As we have discussed before, it is often hard to separate understand statics like chi-square or t-test results if you have too many categories. To help with this, we will only be using three days: one early (Day 3), one in the middle (Day 10), and one at the end (Day 17)

To accomplish this, subset the data frame using the `subset.data.frame()` function to limit the data to only these days. Remember that when writing logical statements, `==` means equal to and the `|` symbol means "or."

```{r bin trials 1,eval=F}
SummaryData<-subset.data.frame()

```

```{r bin trials 2,echo=F}

SummaryData<-subset.data.frame(SummaryData,SummaryData$Day==3 | SummaryData$Day==10 |SummaryData$Day==17)

```

Make sure to look at each trial and be sure it still look good and make sense. If they do, we can move on.

###Creating individual trials
Next we will make a separate data frame for each trial. Here we will also use the `subset.data.frame` function. 
```{r subset trials 1,eval=F}
Trial1<-
Trial2<-
Trial3<-
Trial4<-
```

```{r subset trials 2,echo=F}
Trial1<-subset.data.frame(SummaryData,SummaryData$Trial==1)
Trial2<-subset.data.frame(SummaryData,SummaryData$Trial==2)
Trial3<-subset.data.frame(SummaryData,SummaryData$Trial==3)
Trial4<-subset.data.frame(SummaryData,SummaryData$Trial==4)
```

##Making Plots

Next we will make several plots to help us understand the data. We will start with plotting mean foraging time against day. 

```{r plots 1}
plot(MeanForagingTime~Day,data=Trial1,type='o',ylim=c(0,60),ylab='Foraging Time (s)')
lines(MeanForagingTime~Day,data=Trial2,type='o',col='blue')
lines(MeanForagingTime~Day,data=Trial3,type='o',col='red')
lines(MeanForagingTime~Day,data=Trial4,type='o',col='green')
legend('topleft',legend=c('Trial1','Trial2','Trial3','Trial4'),lty=rep(1,4),pch=rep(1,4),col=c('black','blue','red','green'))
```

Now go ahead and make plots for Percent Starved, Percent Depredated, and Percent Survived

```{r plots2, eval=F,echo=F}
# plot(PStarved~Day,data=Trial1,type='o',ylim=c(0,1),ylab='Percent Starved')
# lines(PStarved~Day,data=Trial2,type='o',col='blue')
# lines(PStarved~Day,data=Trial3,type='o',col='red')
# lines(PStarved~Day,data=Trial4,type='o',col='green')
# legend('topleft',legend=c('Trial1','Trial2','Trial3','Trial4'),lty=rep(1,4),pch=rep(1,4),col=c('black','blue','red','green'))
# 
# plot(PDepredated~Day,data=Trial1,type='o',ylim=c(0,1),ylab='Percent Depredated')
# lines(PDepredated~Day,data=Trial2,type='o',col='blue')
# lines(PDepredated~Day,data=Trial3,type='o',col='red')
# lines(PDepredated~Day,data=Trial4,type='o',col='green')
# legend('topleft',legend=c('Trial1','Trial2','Trial3','Trial4'),lty=rep(1,4),pch=rep(1,4),col=c('black','blue','red','green'))
# 
# plot(PSurvived~Day,data=Trial1,type='o',ylim=c(0,1),ylab='Percent Survived')
# lines(PSurvived~Day,data=Trial2,type='o',col='blue')
# lines(PSurvived~Day,data=Trial3,type='o',col='red')
# lines(PSurvived~Day,data=Trial4,type='o',col='green')
# legend('bottomleft',legend=c('Trial1','Trial2','Trial3','Trial4'),lty=rep(1,4),pch=rep(1,4),col=c('black','blue','red','green'))
```
##Asignment
For the rest of the lab, you will be placed into groups. Each group will be assigned a question. It is up to your group to figure out how to analyze the data. Be sure to check in with me after you have some idea of how to analyze your question to make sure you are doing it right. At the end we will all come together and share our results. Also, I will have you send me your code for your question and compile them to be distributed to the class so everyone can see what you have done.

For your question, you will first need to figure out:
1) What trials will you use?
2) What variables will you test?
3) What statistical test will you use to compare them?
4) What figure can I make to show the results?
5) Make a results statement for your findings. 

One final note, I have provided a function for you to run T-tests with this data if you need it, but you will have to figure out how it works.

```{r t-test function}
TTest<-function(m1,m2,SD1,SD2,N1,N2){
  t<-abs((m1-m2)/sqrt(SD1^2/N2+SD2^2/N1))
  df<-N1+N2-2
  p<-2*pt(t,df,lower.tail=F)
  Res<-data.frame(t,df,p)
  return(Res)
}
```


###Question 1
What was the effect of depleting resources on solitary foraging time?


```{r Quesiton 1 t-test, echo=F}
Test12<-TTest(Trial1$MeanForagingTime[1],Trial1$MeanForagingTime[2],Trial1$StandardDeviation[1],Trial1$StandardDeviation[2],Trial1$N[1],Trial1$N[2])
Test23<-TTest(Trial1$MeanForagingTime[2],Trial1$MeanForagingTime[3],Trial1$StandardDeviation[2],Trial1$StandardDeviation[3],Trial1$N[2],Trial1$N[3])
Test13<-TTest(Trial1$MeanForagingTime[1],Trial1$MeanForagingTime[3],Trial1$StandardDeviation[1],Trial1$StandardDeviation[3],Trial1$N[1],Trial1$N[3])
```



###Question 2
What is the effect of competition in foraging time?

```{r Quesiton 2 worked, echo=F,eval=F}
Test1<-TTest(Trial1$MeanForagingTime[1],Trial2$MeanForagingTime[1],Trial1$StandardDeviation[1],Trial2$StandardDeviation[1],Trial1$N[1],Trial2$N[1])
Test2<-TTest(Trial1$MeanForagingTime[2],Trial2$MeanForagingTime[2],Trial1$StandardDeviation[2],Trial2$StandardDeviation[2],Trial1$N[2],Trial2$N[2])
Test3<-TTest(Trial1$MeanForagingTime[3],Trial2$MeanForagingTime[3],Trial1$StandardDeviation[3],Trial2$StandardDeviation[3],Trial1$N[3],Trial2$N[3])

```

###Question 3
What was the effect of depredation on forager survivorship?

```{r Quesiton 3 worked, echo=F,eval=F}
Obs<-data.frame(Trial1$NSurvived,Trial3$NSurvived)
Test<-chisq.test(Obs)
```

###Question 4
Is flocking beneficial in terms of depredation?

```{r Quesiton 4 worked, echo=F,eval=F}
Obs<-data.frame(Trial1$NSurvived,Trial3$NSurvived)
Test<-chisq.test(Obs)
```

###Question 5
Is flocking beneficial overall?

```{r Quesiton 5 worked, echo=F,eval=F}
Obs<-data.frame(Trial3$NSurvived,Trial4$NSurvived)
Test<-chisq.test(Obs)
```