---
title: "Introduction to R: Basic Programming Concepts: loops, functions, and conditionals"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval=FALSE)
```

Relevant help pages can be found with help("Control") and help("Function").

## For loops
In R a for-loop takes this form,  `variable` is the name of your iteration variable, and `sequence` is a vector or list of values:
```{r}
for (variable in sequence) expression
```
The expression can be a single R command - or several lines of commands wrapped in curly brackets (aka `braces`):

```{r}
for (variable in sequence) { 
    expression
    expression
    expression
}

```

Here is a quick trivial example, printing the square root of the integers one to ten. Copy this code into a script and run to check that it works
```{r}
for (x in c(1:10)) {
  print(sqrt(x))
}
```

## While loops
In R a while takes this form, where condition evaluates to a boolean (True/False) and must be wrapped in ordinary brackets:
```{r}
while (condition) expression
```

As with a for loop, expression can be a single R command - or several lines of commands wrapped in curly brackets:

```{r}
>while (condition) {
    expression
    expression
    expression
}
```

We'll start by using a `while loop` to print out the first few Fibonacci numbers: 0, 1, 1, 2, 3, 5, 8, 13, ... where each number is the sum of the previous two numbers. Create a new R script file, and copy this code into it and run it. 
```{r,eval=FALSE}
a <- 0
b <- 1
print(a)
while (b < 50) {
    print(b)
    temp <- a + b
    a <- b
    b <- temp
}
```

The code works fine, but we would prefer to save the whole sequence in a single vector. This next version builds up the answer gradually using a vector, which it prints at the end:
```{r,eval=TRUE}
x <- c(0,1)   #create a vector with the first two values
while (length(x) < 10) {
    new <- x[length(x)] + x[length(x)-1]
    x <- c(x,new)  #add the new value to the end of the vector
}
print(x)

```


To understand how this manages to append the new value to the end of the vector x, try this at the console prompt:
```{r}
x <- c(1,2,3,4)
c(x,5)
```


## Writing Functions
This following script uses the function() command to create a function (based on the code above) which is then stored as an object with the name Fibonacci:
```{r,eval=TRUE}

Fibonacci <- function(n) {
    x <- c(0,1)
    while (length(x) < n) {
        position <- length(x)
        new <- x[position] + x[position-1]
        x <- c(x,new)
    }
    return(x)
}
```

Once you run this code, there will be a new function available which we can now test:
```{r}
Fibonacci(10)
Fibonacci(3)
Fibonacci(2)
Fibonacci(1)
```


That seems to work nicely - except in the case n = 1 where the function is returning the first two Fibonacci numbers! This gives us an excuse to introduce the if statement.

##  The If statement
In order to fix our function we can do this:
```{r}
Fibonacci <- function(n) {
    if (n==1) return(0)
    x <- c(0,1)
    while (length(x) < n) {
        position <- length(x)
        new <- x[position] + x[position-1]
        x <- c(x,new)
    }
    return(x)
}
```


In the above example we are using the simplest possible if statement:
```{r}
if (condition) expression
```

The if statement can also be used like this:
```{r}
if (condition) expression else expression
```

And, much like the while and for loops the expression can be multiline with curly brackets:

```{r, eval=TRUE}
Fibonacci <- function(n) {
    if (n==1) {
        x <- 0
    } else {
        x <- c(0,1)
        while (length(x) < n) {
            position <- length(x)
            new <- x[position] + x[position-1]
            x <- c(x,new)
        }
    }
    return(x)
}
```

test the new function

```{r}
Fibonacci(1)
Fibonacci(30)
```

